\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in,marginparwidth=2in]{geometry}

% use Unicode characters
\usepackage[utf8]{inputenc}

% clean citations
\usepackage{cite}

% hyperref makes references clicky
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% improves typesetting in LaTeX
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% use adjustwidth environment to exceed text width
\usepackage{changepage}

% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

% remove brackets from references
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% headrule, footrule and page numbers
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}

% colored text
\usepackage{color}

% define custom colors
\definecolor{Gray}{gray}{.25}

% graphics
\usepackage{graphicx}

% side captions
\usepackage{sidecap}

% text wrap around figures
\usepackage{wrapfig}
\usepackage[pscoord]{eso-pic}
\usepackage[fulladjust]{marginnote}
\reversemarginpar

% additional packages for math and tables
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{enumitem}

% ── document begins ──────────────────────────────────────────────────────
\begin{document}
\vspace*{0.35in}

% title
\begin{flushleft}
{\Large
\textbf\newline{\texttt{maxentcpp}: A C++ reimplementation of Maximum Entropy species distribution modeling for R.}
}
\newline
% authors
\\
\'Angel Luis Robles Fern\'andez\textsuperscript{1,*}
\\
\bigskip
\bf{1} Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence, KS, United States
\\
\bigskip
* a.l.robles.fernandez@gmail.com
\\
ORCID: \href{https://orcid.org/0000-0002-4741-8012}{0000-0002-4741-8012}

\end{flushleft}

% ══════════════════════════════════════════════════════════════════════════
\section*{Abstract}
% ══════════════════════════════════════════════════════════════════════════

\texttt{maxentcpp} is an R package offering a native C++17
reimplementation of the Maximum Entropy (Maxent) algorithm for species
distribution modeling~(SDM). The package translates the core
continuous-predictor Maxent~3.4.4 fitting and projection algorithm into
C++17 with R~bindings via Rcpp, eliminating the Java Runtime Environment
dependency entirely. It supports all five feature types, output
transformations in raw, logistic, and cloglog scales, and a streaming
raster evaluation engine for continental-scale projections. Numerical
fidelity is validated per-iteration against the original Java
implementation via a companion test package.

% now start line numbers
\linenumbers

% ══════════════════════════════════════════════════════════════════════════
\section*{Summary}
% ══════════════════════════════════════════════════════════════════════════

\texttt{maxentcpp} is an R package offering a native C++17
reimplementation of the Maximum Entropy (Maxent) algorithm for species
distribution modeling~(SDM).
Maxent estimates the geographic distribution of a species from
presence-only occurrence records and environmental covariates by
identifying the probability distribution of maximum entropy subject to
constraints derived from training data
\cite{Phillips2006,Phillips2004}.
Since its introduction, Maxent and its foundational methodological papers
have been cited extensively across ecology, conservation biology, and
biogeography \cite{Elith2011}.

The fitted model is a \emph{Gibbs distribution} over geographic space:
\begin{equation}\label{eq:gibbs}
  P(\mathbf{x}) \;=\;
    \frac{1}{Z}\,
    \exp\!\Bigl(\sum_{j=1}^{J} \lambda_j\, f_j(\mathbf{x})\Bigr),
\end{equation}
where $f_j$ are feature transformations of the environmental variables,
$\lambda_j$ are learned weights, and
$Z = \sum_{\mathbf{x}}\exp\!\bigl(\sum_j \lambda_j f_j(\mathbf{x})\bigr)$
is the normalizing constant (partition function).
The optimizer finds the $\lambda_j$ that maximize the regularized
log-likelihood via sequential coordinate ascent with
$\ell_1$ (lasso) penalties \cite{Phillips2006,Elith2011}.

The original Maxent software is a Java desktop application
\cite{Phillips2017}.
\texttt{maxentcpp} translates the core continuous-predictor Maxent~3.4.4
fitting and projection algorithm into C++17 with R~bindings via
Rcpp~\cite{Eddelbuettel2011} and RcppEigen~\cite{Bates2013}, eliminating
the Java Runtime Environment dependency entirely.
The package supports all five feature types (linear, quadratic, product,
hinge, and threshold), output transformations in raw, logistic, and
complementary log-log (cloglog) scales, and a streaming raster
evaluation engine that processes environmental layers block-by-block
through the \texttt{terra} package~\cite{Hijmans2024}, enabling
continental-scale projections on commodity hardware without loading
entire raster stacks into memory.
Tools for diagnosing model performance include AUC evaluation, permutation
importance, response curves, and Multivariate Environmental Similarity
Surfaces (MESS) for novelty detection.

% ══════════════════════════════════════════════════════════════════════════
\section*{Statement of need}
% ══════════════════════════════════════════════════════════════════════════

Maxent is the \emph{de facto} standard for presence-only species
distribution modeling~\cite{Merow2013}.
However, the original Java implementation presents a number of practical
barriers for present-day ecological research workflows:

\begin{enumerate}[leftmargin=*,nosep]
  \item \textbf{Java dependency.}
    Java and \texttt{rJava} configuration can be a recurring source of
    installation and runtime failures, particularly across operating
    systems, Java versions, and managed computing environments.
    While \texttt{conda} environments and Docker containers can manage
    Java versioning, they add deployment complexity and are not standard
    practice in many ecology research groups.

  \item \textbf{File-based I/O.}
    Data must be serialized to disk as SWD or ASCII grid files before
    the Java process can read them, and results must be parsed back
    from output files.
    There is no in-memory bridge between R data structures and the Java
    optimizer.

  \item \textbf{Scalability.}
    The Java implementation is single-threaded and GUI-centric.
    In multi-species, multi-scenario workflows (e.g., 500
    species $\times$ 4 climatic scenarios), file~I/O overhead
    accumulates: each species run requires writing SWD files, invoking
    the JVM, and parsing output files.
    A~native in-memory API eliminates this per-call overhead entirely.

  \item \textbf{Maintenance.}
    The Java Maxent codebase has received only minor updates since
    version~3.4.4, with limited active development~\cite{Phillips2017}.
\end{enumerate}

\texttt{maxentcpp} addresses these barriers by providing a native C++/R
implementation.
At present, the development version can be installed from GitHub using
\texttt{remotes::install\_github("alrobles/maxentcpp")}; following CRAN
acceptance, the package will be installable with
\texttt{install.packages("maxentcpp")}.
The package operates entirely in memory through R objects and integrates
seamlessly with the \texttt{terra} spatial data ecosystem.
The intended users are ecologists, conservation biologists, and
biogeographers who employ Maxent in reproducible, script-based
workflows---from individual species analyses to large-scale biodiversity
assessments.

% ══════════════════════════════════════════════════════════════════════════
\section*{Legacy software modernization in ecology}
% ══════════════════════════════════════════════════════════════════════════

\texttt{maxentcpp} fits within a broader modernization trend in
ecological and environmental-science software.
Circuitscape.jl~\cite{Hall2021} illustrates how established ecological
algorithms can be reimplemented in a modern high-performance language to
improve scalability and parallel execution, while retaining continuity
with a widely used landscape-connectivity tool.
In~R, the spatial ecosystem has similarly moved from older
\texttt{sp}/\texttt{raster}-centered workflows toward
\texttt{sf}~\cite{Pebesma2018} and \texttt{terra}~\cite{Hijmans2024},
and SDM packages such as \texttt{ENMeval}~\cite{Kass2021} and
\texttt{biomod2}~\cite{Thuiller2009} have followed this
transition---with \texttt{ENMeval} additionally removing
Java/\texttt{rJava} from its default path in favor of \texttt{maxnet}.

Within Maxent workflows specifically, existing modernization efforts
address different parts of the problem.
\texttt{rmaxent}~\cite{Baumgartner2017} provides Java-free projection of
previously fitted Maxent models from \texttt{.lambdas} files, parses
feature weights and normalizers, and implements MESS-related diagnostics,
but does not replace Java for model fitting.
\texttt{maxnet}~\cite{Phillips2024maxnet} provides a complete Java-free
fitting implementation based on \texttt{glmnet} coordinate
descent---preserving the same statistical model but employing a different
optimization algorithm that can produce numerically different results,
particularly for threshold and hinge features.

\texttt{maxentcpp} occupies a distinct position within this landscape by
porting the Java Maxent \texttt{density.Sequential} training optimizer
itself to C++17 and validating optimizer trajectories against the Java
implementation.
To the best of current knowledge, \texttt{maxentcpp} is the first
R~package to target the Java Maxent~3.4.4 \texttt{density.Sequential}
optimizer through a native C++ port and to publish per-iteration
trajectory tests via the companion package
\texttt{maxentcppCompTest}~\cite{maxentcppCompTest2025}.
Its contribution is not modernization in general, but optimizer-level
fidelity combined with integration into modern R/\texttt{terra}
workflows.

% ══════════════════════════════════════════════════════════════════════════
\section*{State of the field}
% ══════════════════════════════════════════════════════════════════════════

Several R packages provide access to Maxent-family models.
\textbf{dismo}~\cite{Hijmans2023} is the most widely used R interface
to Java Maxent; it wraps \texttt{maxent.jar} via \texttt{rJava},
inheriting the JDK dependency and file~I/O overhead, and is currently in
maintenance mode following the transition from \texttt{raster} to
\texttt{terra}.
\textbf{maxnet}~\cite{Phillips2024maxnet} is a pure-R implementation
that uses \texttt{glmnet} for elastic-net regularization; while it
avoids the Java dependency, it employs a different optimization
algorithm (elastic net via coordinate descent on the full feature
matrix) rather than the sequential coordinate-ascent optimizer of the
original Maxent, which can produce numerically different results,
particularly for threshold and hinge features.
\textbf{ENMeval}~\cite{Kass2021} is a model tuning and evaluation
framework that wraps either \texttt{dismo::maxent()} or
\texttt{maxnet::maxnet()}; it does not implement the Maxent algorithm
itself.
\textbf{kuenm}~\cite{Cobos2019} is a calibration and evaluation
toolkit that relies on \texttt{dismo} or direct Java Maxent calls.

\subsection*{Empirical comparison: maxentcpp vs maxnet}

To quantify the practical differences between \texttt{maxentcpp} and
\texttt{maxnet}, we compared both packages on a virtual
species~\cite{Leroy2016} with known true habitat suitability, constructed
from the environmental layers bundled with \texttt{maxentcpp} (Annual
Mean Temperature and Annual Precipitation over Central America, 2,371
non-NA cells).
The virtual species has a Gaussian niche centered at
20\,\textdegree{}C and 1,500\,mm, and 100 presence records were sampled
proportionally to the true suitability surface (see the companion
vignette \emph{Virtual Species Comparison}\footnote{%
  \url{https://alrobles.github.io/maxentcpp/articles/virtual_species_comparison.html}}
for full reproducible code).

Both packages were fitted with linear, quadratic, and hinge features.
In this illustrative example, the two packages produced highly similar
rank and spatial-overlap patterns:

\begin{table}[!ht]
\centering
\caption{{\bf Ecological equivalence metrics:} \texttt{maxentcpp} vs
         \texttt{maxnet} on a virtual species.}
\begin{tabular}{|l|r|}
\hline
\textbf{Metric} & \textbf{Value} \\
\hline
Schoener's $D$ & 0.93 \\
\hline
Spearman $\rho$ (rank correlation) & 0.95 \\
\hline
Pearson $r$ & 0.97 \\
\hline
Mean $|\Delta\text{cloglog}|$ & 0.053 \\
\hline
Max $|\Delta\text{cloglog}|$ & 0.43 \\
\hline
\end{tabular}
\label{tab:equivalence}
\end{table}

These results suggest that \texttt{maxentcpp} and \texttt{maxnet}
can produce highly similar predictions in a simple continuous-predictor
setting (Schoener's $D > 0.9$ is conventionally interpreted as high
niche overlap), while also showing non-negligible local differences in
the tails of the suitability distribution, where the two optimization
algorithms (\texttt{maxentcpp}: sequential coordinate ascent;
\texttt{maxnet}: elastic-net coordinate descent) regularize differently.
A broader simulation study across multiple species, sample sizes, and
predictor correlation structures would be needed to establish general
equivalence.

The key practical scenarios where a user must choose \texttt{maxentcpp}
over \texttt{maxnet} are:
(1)~when exact reproduction of Java Maxent results is required
(e.g., replicating published analyses),
(2)~when lambda file interoperability with Java Maxent is needed, and
(3)~when streaming raster projection onto grids larger than available
RAM is required.

\subsection*{Performance benchmarks}

Training time for the bundled \emph{Abeillia abeillei} dataset
(73~presences, 2,371~background, linear + quadratic + hinge features):

\begin{table}[!ht]
\centering
\caption{{\bf Training time comparison.}}
\begin{tabular}{|l|r|}
\hline
\textbf{Package} & \textbf{Median time per fit} \\
\hline
\texttt{maxentcpp} & $\sim$820\,ms \\
\hline
\texttt{maxnet}    & $\sim$530\,ms \\
\hline
\end{tabular}
\label{tab:benchmark}
\end{table}

\texttt{maxnet} is faster for small datasets because \texttt{glmnet}'s
coordinate descent is highly optimized for dense feature matrices.
\texttt{maxentcpp}'s streaming evaluation avoids materializing the full
prediction matrix, which is expected to reduce peak memory during
projection; future benchmarks should quantify this advantage on rasters
larger than available RAM.

\texttt{maxentcpp} was built as a new package rather than contributing
to existing projects for three reasons.
First, a faithful port of the original Java optimizer (sequential
coordinate ascent using $\ell_1$-penalized Newton steps) requires a C++
engine that cannot be expressed through \texttt{glmnet}'s API;
contributing this to \texttt{maxnet} would change its core algorithm.
Second, \texttt{dismo}'s architecture is tightly coupled to
\texttt{rJava} and \texttt{raster}; the native C++ strategy eliminates
this dependency chain entirely.
Third, the header-based C++ library design of \texttt{maxentcpp} enables
reuse in other packages or standalone applications beyond~R---something
not achievable through modifications to existing R-only packages.

% ══════════════════════════════════════════════════════════════════════════
\section*{Software design}
% ══════════════════════════════════════════════════════════════════════════

\subsection*{Architecture}

\texttt{maxentcpp} is organized in three layers (C++ core, Rcpp bridge,
R~interface), with the C++ core further split into algorithmic and
infrastructure sublayers:

\begin{table}[!ht]
\begin{adjustwidth}{-1.5in}{0in}
\centering
\caption{{\bf Architecture of \texttt{maxentcpp}:} three layers with the
         C++ core divided into algorithmic and infrastructure sublayers.}
\begin{tabular}{|l|l|r|}
\hline
\textbf{Layer} & \textbf{Key components} & \textbf{Lines} \\
\hline
C++ core (algorithmic)
  & \texttt{Sequential}, \texttt{FeaturedSpace}, five feature types
  & $\sim$4\,500 \\
\hline
C++ core (I/O, diagnostics)
  & \texttt{BackgroundProvider}, grid I/O, CSV, MESS, response curves
  & $\sim$2\,400 \\
\hline
Rcpp bridge
  & \texttt{rcpp\_*.cpp} external-pointer bindings~\cite{Eddelbuettel2013}
  & $\sim$3\,300 \\
\hline
R interface
  & \texttt{maxent\_run()}, feature generators, projection, evaluation,
    diagnostics
  & $\sim$2\,500 \\
\hline
\end{tabular}
\label{tab:arch}
\end{adjustwidth}
\end{table}

The C++ core uses Eigen~\cite{Guennebaud2010} for dense linear algebra.
Of the $\sim$6\,900 C++ lines, approximately 4\,500 implement the core
algorithmic logic (optimizer, features, density) while the remainder
handles I/O and diagnostics.
The high-level \texttt{maxent\_run()} function provides a one-call
workflow mirroring the Java GUI experience, while lower-level functions
(\texttt{maxent\_generate\_features()}, \texttt{maxent\_featured\_space()},
\texttt{maxent\_fit()}, \texttt{maxent\_project\_cloglog()}) offer full
control over each modeling step.

\subsection*{Optimization algorithm}

The \texttt{Sequential} optimizer is a direct port of
\texttt{density.Sequential} in Java Maxent~3.4.4, as represented by the
public \texttt{mrmaxent/Maxent} repository and AMNH release notes (where
3.4.4 is described as a minor bug-fix release).
It minimizes the $\ell_1$-regularized negative log-likelihood:
\begin{equation}\label{eq:objective}
  \mathcal{L}(\boldsymbol{\lambda})
  = -\frac{1}{m}\sum_{i=1}^{m} \sum_{j=1}^{J}
      \lambda_j\, f_j(\mathbf{x}_i)
    + \log Z(\boldsymbol{\lambda})
    + \sum_{j=1}^{J} \beta_j\,|\lambda_j|,
\end{equation}
where $m$ is the number of presence records,
$Z(\boldsymbol{\lambda})
= \sum_{k=1}^{N} \exp\!\bigl(\sum_j \lambda_j f_j(\mathbf{x}_k)\bigr)$
is the partition function summed over $N$ background points, and
$\beta_j = \hat\sigma_j / \sqrt{m}$ is the per-feature regularization
parameter with $\hat\sigma_j$ being the standard deviation of feature~$j$
over the presence points (floored at 0.001).

At each iteration, the optimizer selects the feature~$j^*$ with the most
negative $\Delta\mathcal{L}$ bound (the \texttt{deltaLossBound}
function), then computes a Newton step:
\begin{equation}\label{eq:newton}
  \alpha_j = -\frac{\partial \mathcal{L}/\partial \lambda_j}
    {\mathbf{u}_j^\top \mathbf{H}\, \mathbf{u}_j},
\end{equation}
where $\mathbf{u}_j$ is the $j$-th coordinate direction and
$\mathbf{u}_j^\top \mathbf{H}\, \mathbf{u}_j = \mathrm{Var}_q[f_j]$
is the variance of feature~$j$ under the current distribution~$q$.
The derivative includes the $\ell_1$ subgradient:
$\partial \mathcal{L}/\partial \lambda_j = \mathbb{E}_q[f_j] -
\mathbb{E}_{\hat{p}}[f_j] + \beta_j\,\mathrm{sign}(\lambda_j)$.
The step is damped during early iterations
($\alpha \leftarrow \alpha/50$ for iterations $<10$,
$\alpha/10$ for $<20$, $\alpha/3$ for $<50$) and falls back to
a line search (\texttt{searchAlpha}) if the Newton step does not
decrease loss.
Every 10 iterations, a parallel update applies coordinate steps to all
features simultaneously.

Convergence is tested every 20 iterations: training stops when the
relative loss improvement falls below $10^{-5}$ or 500 iterations are
reached.
All constants in this section are taken directly from Java Maxent~3.4.4
source files, and each statement is covered by a source-mapping test in
\texttt{maxentcppCompTest}.

\subsection*{Streaming raster evaluation}

\texttt{maxentcpp} reads \texttt{terra::SpatRaster} objects
block-by-block through a \texttt{BackgroundProvider} abstraction.
Each tile is scored by the C++ engine and discarded before the next tile
is loaded, allowing projection onto raster stacks larger than available
RAM\@.
The partition function~$Z$ is accumulated across tiles by summing
unnormalized densities
$\exp(\sum_j \lambda_j f_j(\mathbf{x}_k))$ for each cell~$k$ in the
tile, then normalizing once after all tiles are processed.
This produces results equivalent to single-pass evaluation because the
mathematical sum is commutative and each cell is scored independently;
floating-point summation order may introduce differences at the
least-significant bit level.

\subsection*{Numerical fidelity}

Each C++ class is a direct translation of its Java counterpart,
preserving the same control flow, regularization logic, and numerical
constants.
This design choice prioritizes backward compatibility: users migrating
from Java Maxent should obtain numerically equivalent results for
implemented features under matched inputs, defaults, and deterministic
settings.
The fidelity contract is enforced by a dedicated companion package,
\texttt{maxentcppCompTest}~\cite{maxentcppCompTest2025}, which runs the
C++ optimizer and the original Java \texttt{density.Sequential} on
shared test fixtures and compares per-iteration trajectories of loss,
entropy, and $\lambda$~vectors.

On controlled test fixtures (identical input data, same floating-point
representation), agreement reaches $< 10^{-14}$ relative error for
symmetric fixtures and $< 10^{-6}$ on
$\|\Delta\lambda\|_\infty$ for asymmetric fixtures at every iteration
checkpoint.
\textbf{These bounds hold under the specific conditions of the test
fixtures} (same compiler family, IEEE~754 double precision, deterministic
iteration order).
Floating-point ordering differences introduced by alternative BLAS
backends or aggressive compiler optimizations (e.g., \texttt{-ffast-math})
could widen the gap, though the sequential nature of the coordinate-ascent
algorithm limits sensitivity to parallelism-induced reordering.
Cross-platform CI (Ubuntu, macOS, Windows with GCC, Clang, and MSVC)
confirms that the test fixtures pass on all three compiler families.

\begin{table}[!ht]
\centering
\caption{{\bf Class-level porting map} from Java Maxent to
         \texttt{maxentcpp}.}
\begin{tabular}{|l|l|}
\hline
\textbf{\texttt{maxentcpp} (C++)} & \textbf{Java Maxent original} \\
\hline
\texttt{LinearFeature}      & \texttt{density.LinearFeature}    \\
\hline
\texttt{QuadraticFeature}   & \texttt{density.SquareFeature}    \\
\hline
\texttt{ProductFeature}     & \texttt{density.ProductFeature}   \\
\hline
\texttt{HingeFeature}       & \texttt{density.HingeFeature}     \\
\hline
\texttt{ThresholdFeature}   & \texttt{density.ThresholdFeature} \\
\hline
\texttt{FeaturedSpace}      & \texttt{density.FeaturedSpace}    \\
\hline
\texttt{Sequential::run()}  & \texttt{density.Sequential.run()} \\
\hline
\end{tabular}
\label{tab:porting}
\end{table}

\paragraph{Lambda file compatibility.}
Trained models are serialized in the same \texttt{.lambdas} text format
used by Java Maxent~3.4.4.
Lambda files produced by either implementation can be loaded by the
other, ensuring interoperability with existing workflows and archived
models.

\subsection*{Feature completeness}

The following table summarizes the current scope of \texttt{maxentcpp}
relative to Java Maxent~3.4.4 and \texttt{maxnet}.
Legend: $\checkmark$ = implemented and tested;
$\circ$ = partial or experimental;
--- = not implemented.

\begin{table}[!ht]
\begin{adjustwidth}{-1.5in}{0in}
\centering
\caption{{\bf Feature completeness:} \texttt{maxentcpp} vs Java Maxent
         vs \texttt{maxnet}.}
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Feature} & \textbf{\texttt{maxentcpp}} & \textbf{Java Maxent 3.4.4} & \textbf{\texttt{maxnet} 0.1.4} \\
\hline
Linear features              & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
\hline
Quadratic features           & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
\hline
Product features             & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
\hline
Hinge features               & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
\hline
Threshold features           & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
\hline
Raw/logistic/cloglog output  & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
\hline
AUC evaluation               & $\checkmark$ & $\checkmark$ & via ENMeval \\
\hline
Permutation importance       & $\checkmark$ & $\checkmark$ & via ENMeval \\
\hline
Response curves              & $\checkmark$ & $\checkmark$ & $\circ$ \\
\hline
MESS maps                    & $\checkmark$ & $\checkmark$ & --- \\
\hline
Clamping / fade-by-clamping  & $\checkmark$ & $\checkmark$ & --- \\
\hline
Lambda file I/O              & $\checkmark$ & $\checkmark$ & --- \\
\hline
Streaming raster projection  & $\checkmark$ & --- & --- \\
\hline
Bias grids                   & $\checkmark$ & $\checkmark$ & via \texttt{offset} \\
\hline
Categorical variables        & --- & $\checkmark$ & $\checkmark$ \\
\hline
Replicate runs / cross-val.  & --- & $\checkmark$ & via ENMeval \\
\hline
Jackknife variable selection & --- & $\checkmark$ & --- \\
\hline
Missing data handling        & --- & $\checkmark$ & $\checkmark$ \\
\hline
SWD-to-raster projection     & --- & $\checkmark$ & --- \\
\hline
\end{tabular}
\label{tab:features}
\end{adjustwidth}
\end{table}

Categorical variables, replicate runs, jackknife, and SWD-to-raster
projection are not yet implemented.
Users requiring these features should continue using Java Maxent or
\texttt{maxnet}.
We plan to add categorical variable support in a future release.

\subsection*{Development provenance}

The translation followed a phased approach across four publicly
accessible repositories:

\begin{enumerate}[leftmargin=*,nosep]
  \item \texttt{alrobles/Maxent}---a fork of the original
    \texttt{mrmaxent/Maxent}~\cite{Phillips2017} Java source code,
    where the architecture was analyzed and each Java class was mapped to
    a C++ target (193~commits).
  \item \texttt{alrobles/maxentcpp-devel}---the development repository
    where the C++ port, Rcpp bindings, R interface, tests, and
    documentation were iteratively built and refined (34~commits).
  \item \texttt{alrobles/maxentcppCompTest}---the cross-language fidelity
    test suite containing Java oracle runners, golden trajectory files,
    and automated comparison scripts (40~commits).
  \item \texttt{alrobles/maxentcpp}---the release repository with the
    clean, CRAN-ready package (47~commits).
\end{enumerate}

% ══════════════════════════════════════════════════════════════════════════
\section*{Ecosystem comparison}
% ══════════════════════════════════════════════════════════════════════════

The R ecosystem contains several packages that provide access to MaxEnt
models, each employing a distinct integration strategy:

\begin{table}[!ht]
\begin{adjustwidth}{-1.5in}{0in}
\centering
\caption{{\bf R packages providing MaxEnt access.}}
\begin{tabular}{|l|l|l|}
\hline
\textbf{Package} & \textbf{MaxEnt integration} & \textbf{Dependency} \\
\hline
\textbf{dismo}~\cite{Hijmans2023}
  & rJava bridge to \texttt{maxent.jar} & Java JDK, rJava \\
\hline
\textbf{kuenm}~\cite{Cobos2019}
  & \texttt{system2("java -jar maxent.jar")} & Java JDK \\
\hline
\textbf{kuenm2} (Cobos et al.)
  & \texttt{glmnet} via forked \texttt{maxnet} code & glmnet \\
\hline
\textbf{wallace}~\cite{Kass2018wallace}
  & Delegates to ENMeval $\to$ maxnet or dismo & ENMeval \\
\hline
\textbf{ENMTools}~\cite{Warren2010}
  & \texttt{dismo::maxent()} directly & dismo, rJava \\
\hline
\textbf{biomod2}~\cite{Thuiller2009}
  & Java (\texttt{system2}) or \texttt{maxnet} & Optional Java or maxnet \\
\hline
\textbf{rmaxent}~\cite{Baumgartner2017}
  & Java-free projection of fitted Maxent models, \texttt{.lambdas} parsing, MESS & None (projection only) \\
\hline
\textbf{MIAmaxent}~\cite{Vollering2019}
  & MaxEnt via subset selection (not lasso) & glmnet \\
\hline
\textbf{SDMtune}~\cite{Vignali2020}
  & Trains/tunes SDMs via dismo or maxnet & dismo or maxnet \\
\hline
\textbf{maxentcpp}
  & Native C++17 via Rcpp & None (self-contained) \\
\hline
\end{tabular}
\label{tab:ecosystem}
\end{adjustwidth}
\end{table}

These packages can be grouped by their relationship to the MaxEnt
algorithm:

\begin{itemize}[nosep,leftmargin=*]
  \item \textbf{Black-box wrappers} (dismo, kuenm, ENMTools, biomod2
    MAXENT mode): invoke the Java binary and read output files.
    The optimizer is inaccessible.
  \item \textbf{Approximate reimplementations} (maxnet, kuenm2, biomod2
    MAXNET mode): use \texttt{glmnet} coordinate descent on the logistic
    regression formulation.
    The statistical model is equivalent but the optimization path
    differs, which can produce numerically different results for
    threshold and hinge features.
  \item \textbf{Faithful reimplementation} (maxentcpp): ports the
    original Sequential optimizer to C++ and validates per-iteration
    numerical equivalence.
\end{itemize}

The distinguishing contribution of \texttt{maxentcpp} is that it is the
only package offering a compiled native reimplementation of the actual
MaxEnt \texttt{density.Sequential} optimizer---preserving both the
statistical model and the optimization algorithm while eliminating the
Java dependency.

% ══════════════════════════════════════════════════════════════════════════
\section*{Limitations and inherited assumptions}
% ══════════════════════════════════════════════════════════════════════════

The Maxent~3.4.4 algorithm that \texttt{maxentcpp} faithfully reproduces
has well-documented limitations that users should be aware of:

\begin{itemize}[nosep,leftmargin=*]
  \item \textbf{Feature explosion.}
    The number of features (particularly hinge and threshold) grows with
    the number of environmental variables, which can cause
    identifiability issues in high-dimensional settings
    \cite{Elith2011,Merow2013}.

  \item \textbf{Sensitivity to background sampling.}
    Maxent's output is influenced by the choice and size of the
    background sample, which affects the estimated density and can bias
    predictions in undersampled regions~\cite{Phillips2009}.

  \item \textbf{$\ell_1$ regularization limitations.}
    The sequential coordinate ascent with $\ell_1$ penalties produces
    sparse models but does not guarantee selection of the ``correct''
    features under high correlation among predictors~\cite{Hastie2015}.

  \item \textbf{No principled uncertainty quantification.}
    Unlike Bayesian approaches or bootstrap-based methods, the Maxent
    optimizer produces point estimates without confidence intervals.
\end{itemize}

A principled alternative would be to reformulate the Maxent objective
using modern convex optimization tooling (e.g., proximal gradient
methods, ADMM) or Bayesian inference.
However, such reformulations would produce different results than the
widely used Java implementation, breaking comparability with published
analyses.
The design of \texttt{maxentcpp} explicitly aims to preserve this
comparability.

% ══════════════════════════════════════════════════════════════════════════
\section*{CRAN compliance and software quality}
% ══════════════════════════════════════════════════════════════════════════

Achieving CRAN acceptance requires more than passing \texttt{R CMD check}:
packages must meet strict software-quality standards that safeguard user
environments and ensure reproducibility across
platforms~\cite{CRANPolicy2024,WRE2024}.
During the pre-submission review process, \texttt{maxentcpp} underwent
targeted improvements to satisfy these requirements, detailed below.

\subsection*{Example documentation: \texttt{\textbackslash dontrun\{\}}
to \texttt{\textbackslash donttest\{\}}}

CRAN policy distinguishes between examples that \emph{cannot} execute
(missing external software, API keys) and those that are merely slow or
resource-intensive.
The former require \texttt{\textbackslash dontrun\{\}}; the latter must
use \texttt{\textbackslash donttest\{\}} so that CRAN's check
infrastructure can optionally execute them with
\texttt{-{}-run-donttest}~\cite{CRANcookbook2024}.
All 36 example blocks in \texttt{maxentcpp} were originally wrapped in
\texttt{\textbackslash dontrun\{\}}, which suppresses execution entirely
and misleads users with a \texttt{\# Not run:} comment.
These were migrated to \texttt{\textbackslash donttest\{\}}, converting
the documentation examples into live, testable code that CRAN
infrastructure can verify on every platform.

\subsection*{Self-contained examples with synthetic data}

Changing the wrapper from \texttt{\textbackslash dontrun\{\}} to
\texttt{\textbackslash donttest\{\}} revealed a deeper issue: most
examples referenced undefined objects (e.g., a pre-fitted model
\texttt{m}, a grid \texttt{g1}, or a results data frame) that only
existed in the context of a vignette narrative.
Under \texttt{\textbackslash dontrun\{\}} this was invisible because
the code was never evaluated; under \texttt{\textbackslash donttest\{\}}
it caused immediate failures.

Every example was rewritten to be self-contained, generating synthetic
data at the point of use:

\begin{verbatim}
# Example: maxent_permutation_importance
n   <- 50
env <- data.frame(bio1 = runif(n, 10, 30),
                  bio12 = runif(n, 500, 2500))
f   <- maxent_generate_features(env, features = "lqh")
fs  <- maxent_featured_space(f, sample(n, 10))
m   <- maxent_fit(fs)
maxent_permutation_importance(m, fs)
\end{verbatim}

This pattern---using \texttt{runif()}, \texttt{data.frame()}, and the
package's own constructors (\texttt{maxent\_grid\_from\_matrix},
\texttt{maxent\_featured\_space},
\texttt{maxent\_fit})---ensures that each example runs in isolation
without relying on objects from other functions or prior sessions.

\subsection*{Graphics state restoration via \texttt{on.exit()}}

CRAN requires that functions never leave the user's session in an
altered state: any modifications to \texttt{par()}, \texttt{options()},
or the working directory must be reverted, even if the function errors
mid-execution~\cite{WRE2024}.
The \texttt{maxent\_plot\_response\_curves()} function modified
\texttt{par()} inside a loop to configure PNG output files but restored
settings manually at the end of each iteration---a pattern that leaks
state on error.

The fix extracts device management into an internal helper,
\texttt{.maxent\_safe\_png()}, that encapsulates the
open-device~$\to$~set-par~$\to$~plot~$\to$~close-device lifecycle with
guaranteed cleanup:

\begin{verbatim}
.maxent_safe_png <- function(file, width, height, expr) {
  grDevices::png(file, width = width, height = height)
  oldpar <- graphics::par(mar = c(4, 4, 2, 1))
  on.exit({ graphics::par(oldpar); grDevices::dev.off() },
          add = TRUE)
  force(expr)
}
\end{verbatim}

Because \texttt{on.exit()} is function-scoped, placing the cleanup
inside a dedicated helper avoids the anti-pattern of accumulating or
overwriting exit handlers within a loop.

\subsection*{Optional dependency management}

\texttt{maxentcpp} lists \texttt{terra} as a suggested (optional)
dependency for raster I/O operations.
CRAN checks may run with \texttt{\_R\_CHECK\_FORCE\_SUGGESTS\_=false},
causing examples that assume \texttt{terra} is installed to fail.
All terra-dependent examples are now guarded:

\begin{verbatim}
if (requireNamespace("terra", quietly = TRUE)) {
  # example code using terra
}
\end{verbatim}

This ensures graceful skipping when the optional dependency is
unavailable, following CRAN's recommended pattern for structuring
examples that depend on suggested packages~\cite{CRANcookbook2024}.

\subsection*{Continuous integration}

GitHub Actions workflows run \texttt{R CMD check -{}-as-cran} on Ubuntu
(R~release and R-devel), macOS (R~release), and Windows (R~release),
with the \texttt{-{}-run-donttest} flag enabled to exercise all example
code.
The CI matrix also includes a dedicated CRAN-preflight job that sets
\texttt{\_R\_CHECK\_FORCE\_SUGGESTS\_=false} to simulate CRAN's
minimal-dependency environment.

% ══════════════════════════════════════════════════════════════════════════
\section*{Research impact statement}
% ══════════════════════════════════════════════════════════════════════════

\texttt{maxentcpp} currently provides a numerically faithful
implementation of the core continuous-predictor Maxent fitting and
projection workflow for the implemented feature classes and output
transformations.
It is not yet a complete replacement for all Java Maxent workflows,
because categorical predictors, replicate runs, jackknife diagnostics,
missing-data handling, and SWD-to-raster projection are not implemented
in the current release.
For those features, users should continue using Java Maxent or
\texttt{maxnet}.

The package is distributed under the MIT license (compatible with
Eigen's MPL2 and RcppEigen's GPL-2 via the ``or later'' clause), with
full source code, a pkgdown documentation website
(\url{https://alrobles.github.io/maxentcpp/}), and four vignettes
covering a getting-started walkthrough, the mathematical foundations of
Maxent features, a comparative motivation document, and a virtual
species validation study.
Continuous integration via GitHub Actions runs \texttt{R CMD check} on
Ubuntu, macOS, and Windows across R~release and R-devel.
The test suite comprises over 20~test files ($\sim$4\,800~lines)
covering feature generation, optimization convergence, spatial
projection, Java numerical equivalency, model diagnostics, and
end-to-end workflows.

By providing a dependency-free, memory-efficient, and numerically faithful
Maxent implementation, \texttt{maxentcpp} may lower the barrier for
large-scale biodiversity assessments, climate-change projections, and
conservation prioritization studies that rely on presence-only species
distribution models, once the remaining Java Maxent workflow
features---especially categorical predictors and replicate
diagnostics---are implemented or clearly documented as out of scope.

% ══════════════════════════════════════════════════════════════════════════
\section*{AI usage disclosure}
% ══════════════════════════════════════════════════════════════════════════

Generative AI tools were used during the development of
\texttt{maxentcpp} and the preparation of this manuscript, as detailed
below.

\paragraph{Tools used.}
\begin{itemize}[leftmargin=*,nosep]
  \item \textbf{GitHub Copilot} (GPT-4-based, 2025 version): code
    scaffolding and boilerplate generation during the initial porting
    phases in \texttt{alrobles/Maxent} and
    \texttt{alrobles/maxentcpp-devel}.
  \item \textbf{Devin} (Cognition AI, 2025--2026): incremental feature
    implementation, test writing, documentation generation, CRAN
    compliance fixes, CI configuration, and manuscript drafting.
  \item \textbf{Claude} (Anthropic, 2025--2026): complex refactoring,
    cross-cutting documentation, and code review.
\end{itemize}

\paragraph{Nature and scope of assistance.}
AI tools contributed to: C++ code generation and refactoring from Java
source, Rcpp binding scaffolding, R wrapper functions, \texttt{testthat}
test scaffolding and expansion, \texttt{roxygen2} documentation,
vignette drafting, pkgdown site configuration, CI workflow setup, CRAN
compliance review, and manuscript drafting.
The core algorithmic C++ classes (\texttt{Sequential},
\texttt{FeaturedSpace}, feature types) were translated from Java with AI
assistance but under direct human supervision: each class was mapped
line-by-line from the corresponding Java source, reviewed for
correctness against the Java reference, and validated through the
\texttt{maxentcppCompTest} trajectory comparison framework.
Approximately 60\% of the C++ algorithmic lines were initially drafted
by AI tools and subsequently reviewed and corrected by the human author;
the remaining 40\% were written directly by the author, particularly
the performance-critical optimizer inner loops and numerical edge cases.

\paragraph{Maintainability.}
The human author (\'A.\,L.\ Robles~Fern\'andez) can independently
explain and modify every algorithmically significant class.
As evidence: the \texttt{Sequential::run()} optimizer,
\texttt{good\_alpha()}, \texttt{delta\_loss\_bound()},
\texttt{newton\_step\_feature()}, and \texttt{do\_parallel\_update()}
methods are documented with line-by-line references to the corresponding
Java source (e.g., ``mirrors Sequential.java:294..342''), and the author
has independently debugged numerical discrepancies during the fidelity
testing process.

\paragraph{Confirmation of review.}
All AI-generated code, tests, documentation, and manuscript text were
reviewed, edited, and validated by the human author
(\'A.\,L.\ Robles~Fern\'andez), who made all core design decisions
including the algorithm translation strategy, API design, numerical
fidelity requirements, and the phased development architecture.
The full development history is publicly available in the four
repositories listed above.

% ══════════════════════════════════════════════════════════════════════════
\section*{Acknowledgements}
% ══════════════════════════════════════════════════════════════════════════

The author thanks Steven~J. Phillips, Robert~P. Anderson, and
Robert~E. Schapire for creating the original Maxent software and making
the source code publicly available under an MIT license.
The \texttt{Rcpp}, \texttt{RcppEigen}, and \texttt{terra} package
maintainers are gratefully acknowledged for providing the infrastructure
that makes \texttt{maxentcpp} possible.

\nolinenumbers

% ══════════════════════════════════════════════════════════════════════════
\bibliography{paper}
\bibliographystyle{abbrv}
% ══════════════════════════════════════════════════════════════════════════

\end{document}
